---
title: "R Ladies Superheat Tutorial"
author: "Rebecca Barter"
date: "9/20/2017"
output: html_document
---

In this tutorial, you will learn how to use superheat, and will discover a portion of the immense variety of customization options.

The superheat vignette can be found at https://rlbarter.github.io/superheat/. This incredibly detailed vignette was created using bookdown and is hosted on github pages and it contains (almost) everything you might want to know about using superheat!

# Setup

First, let's load the packages we will be using throughout the tutorial.

```{r, message=FALSE, warning=FALSE}
# load packages
library(tidyverse)
library(devtools)
library(knitr)
library(forcats)
```

Next, we will install the superheat package from GitHub (using the `install_github()` function from the `devtools` package). Note that superheat is also on CRAN, but this is an older version; the most recent version will always be the GitHub development version (but it might have a few minor bugs here and there!).


```{r, eval = FALSE}
# install superheat package from github
install_github("rlbarter/superheat")
```

```{r}
# install superheat
library(superheat)
```

# Get the data

We will be using data on organ donations from 2006 to 2014 is found in the `organ.csv` file in the data folder. Let's load it in:

```{r}
# load in the organ donation data
organs <- read.csv("data/organs.csv")
colnames(organs) <- c("country", 2006:2014)
# view first 6 rows of the organ donation data
kable(head(organs), digits = 1)
```


# Heatmaps using ggplot2

To make a heatmap in ggplot2, you need to convert your dataframe to long-form.


```{r}
# convert organs to longform using tidyr
organs_long <- organs %>% 
  gather(key = "year", value = "donors", -country)
# look at the first 6 rows of the long-form dataset
kable(head(organs_long), digits = 1)
# look at the last 6 rows of the long-form dataset
kable(tail(organs_long), digits = 1)
```

I could use ggplot to create a heatmap of the organ donations by country. 

```{r, fig.height = 8}
ggplot(organs_long) + 
  geom_raster(aes(x = year, y = country, fill = donors)) +
  scale_fill_viridis_c()
```

Clearly some row ordering is needed! For example, perhaps we want to order the rows in decreasing order. However, while it is easy to rearrange the rows of a matrix when it is stored in its original form, manipulating a matrix when it is recorded in long-form can be surprisingly difficult.

One way to do it is as follows:

```{r}
organs_long <- organs_long %>% 
  # identify the average number of donations per country
  group_by(country) %>%
  mutate(avg_donors = mean(donors, na.rm = T)) %>%
  ungroup() %>%
  arrange(avg_donors) %>%
  # remove the avg_donors column
  select(-avg_donors) %>%
  # reorder the country factor levels
  mutate(country = fct_inorder(country))

```


While this does the job, it is somewhat convoluted, especially for someone who is not a highly experienced R user! The re-ordered plot is shown below:


```{r, fig.height = 8}
ggplot(organs_long) + 
  geom_raster(aes(x = year, y = country, fill = donors)) +
  scale_fill_viridis_c()
```


# Heatmaps using superheat

When using superheat, there is no need to convert the original matrix to a longform.

```{r}
# convert the data to a numeric-only data frame
organs_matrix <- organs
# replace the rownames with the country names
rownames(organs_matrix) <- organs$country
organs_matrix <- organs_matrix %>% select(-country)
```

```{r, fig.height = 14, fig.width = 9}
superheat(organs_matrix)
```

Rearranging the rows is much simpler when dealing with a wideform matrix rather than a longform data frame.

```{r, eval = FALSE}
# identify the order of the rows (increasing order of average donations)
row_order <- apply(organs_matrix, 1, mean, na.rm = T) %>% order
# reorder the rows of the matrix
organs_matrix <- organs_matrix[row_order, ]
```


However, with superheat, not even this is necessary. You can simply provide a row order argument to the function.


```{r, fig.height = 14, fig.width = 9}
superheat(organs_matrix, 
          title = "Number of organs donated by deceased donors\nper 100,000 individuals",
          # arrange the rows in order of increasing average number of donors
          order.rows = order(apply(organs_matrix, 1, mean, na.rm = T)))
```

# The power of superheat: adding additional information


Much of the power of the superheat package comes from the ability to add additional information to the heatmap.




## Adding adjacent variables

You can add additional variables to the plot via adjacent scatter, line, bar, or boxplots. For adjacent plots above the heatmap, the x-axis corresponds to the column variables. Correspondingly, for ajacent plots to the right of the heatmap, the y-axis corresponds to the row variables.


In the superheatmap below, we add a line plot above the columns which corresponds to the total number of organs (per 100,000) donated over time (summing over the countries).


```{r, fig.height = 17, fig.width = 9}
# adding a trendline above the heatmap
superheat(organs_matrix,
          title = "Number of organs donated by deceased donors\nper 100,000 individuals",
          # arrange the rows in order of increasing average number of donors
          order.rows = order(apply(organs_matrix, 1, mean, na.rm = T)),
          
          # add a plot of total organs donated accross time
          yt = apply(organs_matrix, 2, sum, na.rm = T),
          yt.axis.name = "Total organs\nfrom deceased donors",
          yt.plot.type = "line",
          yt.plot.size = 0.25,
          yt.axis.name.size = 12)
```


Next, we can also add external information, such as the human development index (HDI) ranking for each country, as a barplot to the right of the rows.

```{r}
hdi <- read.csv("data/hdi_2014.csv") 
kable(head(hdi))
```

Note that the `order.rows` argument will apply the same ordering to `yr` as to the rows of the matrix.


```{r, fig.height = 17, fig.width = 12}
# add hdi as a barplot to the rows
superheat(organs_matrix,
          title = "Number of organs donated by deceased donors\nper 100,000 individuals",
          # arrange the rows in order of increasing average number of donors
          order.rows = order(apply(organs_matrix, 1, mean, na.rm = T)),
          
          # add a plot of total organs donated accross time
          yt = apply(organs_matrix, 2, sum, na.rm = T),
          yt.axis.name = "Total organs\nfrom deceased donors",
          yt.plot.type = "line",
          yt.plot.size = 0.25,
          yt.axis.name.size = 12,
          
          # add a hdi barplot
          yr = hdi$rank,
          yr.plot.type = "bar",
          yr.axis.name = "HDI ranking",
          yr.axis.name.size = 12)
```

### Perfecting the aesthetics

Having added a bunch of information to our superheatmap, we are now ready to perfect our superheatmap. For instance, we can change the color of each 

```{r, fig.height = 17, fig.width = 12}
# doing a bunch of stuff to make the plot prettier
superheat(organs_matrix,
          title = "Number of organs donated by deceased donors\nper 100,000 individuals",
          # arrange the rows in order of increasing average number of donors
          order.rows = order(apply(organs_matrix, 1, mean, na.rm = T)),
          
          # add a plot of total organs donated accross time
          yt = apply(organs_matrix, 2, sum, na.rm = T),
          yt.axis.name = "Total organs\nfrom deceased donors",
          yt.plot.type = "line",
          yt.plot.size = 0.25,
          yt.axis.name.size = 14,
          
          # add a hdi barplot
          yr = hdi$rank,
          yr.plot.type = "bar",
          yr.axis.name = "HDI ranking",
          yr.axis.name.size = 14,
          yr.obs.col = rep("grey80", nrow(organs_matrix)),
          
          # bottom labels
          bottom.label.size = 0.05,
          bottom.label.col = "white",
          bottom.label.text.angle = 90,
          bottom.label.text.alignment = "right",
          
          # left labels
          left.label.col = "white",
          left.label.text.alignment = "right",
          
          # grid lines
          grid.vline.col = "white",
          grid.vline.size = 2,
          force.grid.hline = TRUE,
          grid.hline.col = "white",
          grid.hline.size = 0.5)
```


## Bonus: Adding text or numbers

As a bonus, if you so desired, you could add the raw counts to the heatmap.

```{r, fig.height = 16, fig.width = 9}
# remove NAs for matrix to plot ontop of heatmap
organs_text <- round(organs_matrix, 1)
organs_text[is.na(organs_text)] <- 0
organs_text <- as.matrix(organs_text)
# plot the matrix on top of the heatmap
superheat(organs_matrix,
          title = "Number of organs donated by deceased donors\nper 100,000 individuals",
          X.text = organs_text)
```

You could also change the color of the text so that the text for darker cells is lighter.

```{r, fig.height = 16, fig.width = 9}
# set text color
organs_text_color <- organs_text
organs_text_color[organs_text < 9] <- "grey80"
organs_text_color[organs_text >= 9] <- "black"
# plot the matrix on top of the heatmap
superheat(organs_matrix,
          title = "Number of organs donated by deceased donors\nper 100,000 individuals",
          X.text = organs_text,
          X.text.col = organs_text_color)
```
